---
title: "WTI Options Portfolio: Full Greek P&L Attribution"
subtitle: "Individual Project"
date: "`r Sys.Date()`"
author: "Your Name"
format:
  html:
    theme: cosmo
    embed-resources: true
    page-layout: full
    toc: true
    toc-location: left
    grid:
      margin-width: 200px
      body-width: 1000px
    resource:
      - "*.js"
      - "*.mp3"
editor: source
execute: 
  echo: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
library(tidyverse)
library(gt)
library(RTL)
library(gridExtra)

F_0  <- 73.00
r    <- 0.04
mult <- 1000

port <- dplyr::tribble(
  ~pos, ~type,   ~K,     ~T2M,   ~sigma,  ~n,
  "P1", "put",   70.00,  3/12,   0.3500,  +2000L,
  "P2", "put",   55.00,  3/12,   0.4200,  -2000L,
  "P3", "call",  80.00,  3/12,   0.3000,  -2000L
)
```

# Description

You are a risk analyst at an energy trading desk. The desk holds a portfolio of WTI crude oil options — specifically, options on NYMEX CL futures contracts.

Your role is to:

+ Build a full-Greek P&L attribution framework: the standard tool used on trading desks to explain daily P&L movements across **Delta** (directional exposure), **Gamma** (curvature), **Vega** (volatility sensitivity), **Vanna** (sensitivity of delta to volatility), and **Theta** (time decay).
+ You are expected to understand contracts and use the right option model and specifications in the `RTL` package.
+ Tell me in your own words the portfolio's risk profile, how the Greeks interact, and what this means for a desk managing this book in practice.
+ No AI use in any shape or form and assignment submitted should be a github repo that I can clone and assess.

# Objective

1. Compute numerical Greeks for each position and for the overall portfolio across the full WTI price range ($50–$100/bbl).
2. Interpret the portfolio's mark-to-market profile and explain how changing Gamma across strike levels affects delta-hedging in practice.
3. Build a complete P&L attribution framework — Delta, Gamma, Vega, Vanna, and Theta — and verify it accurately explains overnight P&L under a realistic market move.
4. Critically evaluate where and why the linear/quadratic approximation breaks down as spot moves away from current market.

# Exposure

```{r exposure_table, echo=FALSE}
dplyr::tribble(
  ~Position, ~Type,  ~`Strike K ($/bbl)`, ~Maturity, ~`Impl. Vol`, ~Contracts,
  "P1", "Put",  70.00, "3M", "35.00%", "+2,000 long",
  "P2", "Put",  55.00, "3M", "42.00%", "-2,000 short",
  "P3", "Call", 80.00, "3M", "30.00%", "-2,000 short"
) |>
  gt::gt() |>
  gt::tab_header(
    title    = "WTI Options Portfolio",
    subtitle = "Three-way collar: long put spread + short OTM call  |  F\u2080 = $73/bbl"
  ) |>
  gt::fmt_number(columns = `Strike K ($/bbl)`, decimals = 2)
```

**Market parameters**: current WTI futures price F₀ = **$73.00/bbl**, risk-free rate r = **4.00%** (annualised, continuously compounded).

Select the appropriate pricing model from the `RTL` package for options on WTI crude oil futures. Consider the nature of the underlying contract — a NYMEX CL futures — when choosing the model and setting its cost-of-carry parameter.

# Question 1 (9 points)

For **each position**, compute the following Greeks **numerically** using central differences evaluated at `F_0`:

+ **Delta** (∂V/∂F)
+ **Gamma** (∂²V/∂F²)
+ **Vega** (∂V/∂σ)
+ **Theta** (∂V/∂T)

**Step size selection**: For each Greek, choose an appropriate finite-difference step size and explain your reasoning. A good step is small enough that the finite difference accurately approximates the true derivative, but not so small that floating-point subtraction errors dominate. Consider the natural scale of each risk factor (price in $/bbl, vol as a decimal, time in years) and any special considerations for the short-dated options in this portfolio.

----

**Answer**

+ **Delta**: Because Delta measure the option's sensitivity to the underlying CL future, and the minimum price fluctuation per the CME is 1 cent per barrel, then the step size should be at least 1 cent. But, to avoid subtraction errors, I checked the listed CL futures options and found that they have strike prices that vary in 50 cent increments I will choose the step size to be 25 cents per barrel. The idea is that If a future moves 25 cents up or down, then an option will either move deeper ITM, deeper OTM, or close to ITM. This will hopefully be a sufficient step size that will not smooth the curve in areas where delta is high and sensitive to price moves.  

+ **Gamma**: Follows the same logic and step size as gamma

+ **Vega**: The options in this portfolio all expire in 3 months, meaning that they are on a steeper end of the volatility smile. That being said, the calls and puts have very different implied vol - the puts have a higher IV so that means that the market values a steep drop in price more than a sharp increase. I would assume that the vegas would also be differnet and that the portfolio vega would be largely drive by the puts as they currently stand, so I want to explore having the step size scale with IV while remaining proportional across the positions. (i.e wih a higher IV we can use a higher step size to capture more of the trend of the smile at that point. So we use 3% * IV as the step size). For now, I will use a 1% step size in IV across the portfolio.

+ **Theta**: For Theta, becasue this measures the sensitivity to passage of time, I will use 1 calendar day as the step size instead of using 1 trading day. Becuse even if the markets are closed, time still goes by and T2M decreases, so I believe that 1 calendar day is an adequate time step. 

-----

**Dollar scaling**: Each raw numerical derivative gives the option price change **per barrel** for a unit move in the risk factor. To express the Greek as a total dollar P&L impact for a position, multiply by `n × mult`, where `n` is the signed contract count (positive = long, negative = short) and `mult = 1,000 bbl/contract`. For example, a delta of 0.50/bbl for a 3,000-contract long position means the desk gains or loses 0.50 × 3,000 × 1,000 = $1,500,000 per $1/bbl move in WTI.

Create plots showing each Greek vs. futures price over F ∈ [50, 100], with one line per position.

**d.** Compute the **portfolio mark-to-market (MTM)** value and **portfolio Gamma** across F ∈ [50, 100]. Plot them on two separate panels. In 3–5 sentences, explain:

- What the MTM profile shape reveals about the portfolio's net directionality and curvature.
- Where portfolio Gamma is highest and lowest across the price range, and why those regions correspond to specific positions.
- What the changing Gamma means for a delta-hedger rebalancing daily as WTI moves through the $50–$100 range.

```{r echo=FALSE}
library(purrr)
library(patchwork)

```


```{r q1, echo=FALSE}


t1 <- dplyr::tribble(
  ~Position, ~Type,  ~`Strike K ($/bbl)`, ~Maturity, ~`Impl. Vol`, ~Contracts,
  "P1", "Put",  70.00, "3M", "35.00%", "+2,000 long",
  "P2", "Put",  55.00, "3M", "42.00%", "-2,000 short",
  "P3", "Call", 80.00, "3M", "30.00%", "-2,000 short")

F0 <- 73
r <- 0.04
T2M <- 3/12
mult <- 1000

delta_step <- 0.25
vega_step <- 1/100
theta_step <- 1/365

# I am assuming these options are American style options, so I will use the CRR model to value them and set the cost of carry equal to 0 because neither the futures nor the options pay dividends, and because the options are on WTI futures, there are also no storage costs related to owning a position in a future's contract, there is however, a financing cost in the form of margin, but I will ignore that for the purposes of this project because the pricing models that we use to value the options are based on a risk neutral probability. 


calculate_crr <- function(S, X, T2M, r, b, N, sigma, type, optionStyle = "american") {
   out <- RTL::CRROption(S, X, sigma, r, b, T2M, N, type, optionStyle)$price
  return(out)}

t1 <- t1 %>% 
  dplyr::mutate(sigma = as.numeric(str_extract(`Impl. Vol`,pattern = "\\d{2}\\.\\d{2}")) / 100,
                X = `Strike K ($/bbl)`,
                Type = tolower(Type),
                n_contracts = dplyr::case_match(
                  Position,
                "P1" ~  2000,
                "P2" ~ -2000,
                "P3" ~ -2000)) %>% 
  dplyr::mutate(
    P_current = purrr::pmap_dbl(
      .l = list(S = F0, X = X,T2M = T2M, r = r, b = 0, N = 1000, sigma = sigma, type = Type, optionStyle = "american"),
      .f = calculate_crr
    ),
    P_up = purrr::pmap_dbl(
      .l = list(S = F0 + delta_step, X = X,T2M = T2M, r = r, b = 0, N = 1000, sigma = sigma, type = Type, optionStyle = "american"),
      .f = calculate_crr
    ),
    P_down = purrr::pmap_dbl(
      .l = list(S = F0 - delta_step, X = X,T2M = T2M, r = r, b = 0, N = 1000, sigma = sigma, type = Type, optionStyle = "american"),
      .f = calculate_crr
    ),
    P_up_vega = purrr::pmap_dbl(
      .l = list(S = F0, X = X,T2M = T2M, r = r, b = 0, N = 1000, sigma = sigma + vega_step, type = Type, optionStyle = "american"),
      .f = calculate_crr
    ) ,
      P_down_vega = purrr::pmap_dbl(
      .l = list(S = F0, X = X,T2M = T2M, r = r, b = 0, N = 1000, sigma = sigma - vega_step, type = Type, optionStyle = "american"),
      .f = calculate_crr
    ),
     P_down_theta = purrr::pmap_dbl(
      .l = list(S = F0, X = X,T2M = T2M - theta_step, r = r, b = 0, N = 1000, sigma = sigma, type = Type, optionStyle = "american"),
      .f = calculate_crr
    ),
    delta = ((P_up - P_down) / (2 * delta_step)) * n_contracts * mult,
    gamma = ((P_up - 2 * P_current + P_down) / (delta_step ^ 2) )* n_contracts * mult,
    vega = ((P_up_vega - P_down_vega) / (2 * vega_step)) * n_contracts * mult,
    theta = (P_down_theta - P_current) * n_contracts * mult
    )


F.grid <- seq(from = 50, to = 100, by  = 0.5)

greeks_df <- t1 %>% 
  dplyr::select(Position,Type,sigma,X, n_contracts) %>% 
  tidyr::crossing(S = F.grid) %>% 
  rowwise() %>% 
  dplyr::mutate(
    results = list(RTL::GBSOption(S=S, X=X, T2M=T2M, r=r, 
                                      b=0, sigma=sigma, type=Type)),
    raw_delta   = results$delta,
    raw_gamma   = results$gamma,
    raw_vega    = results$vega,
    raw_theta   = results$theta,
    
    delta       = raw_delta * n_contracts * 1000,
    gamma       = raw_gamma * n_contracts * 1000,
    vega        = raw_vega  * n_contracts * 1000,
    theta       = raw_theta * n_contracts * 1000) %>% 
  
  dplyr::ungroup() %>% 
  dplyr::select(-results,-starts_with("raw")) %>% 
  dplyr::mutate(Position = case_match(
    Position,
    "P1" ~ "Long Put  X=$70 σ=35%",
    "P2" ~ "Short Put  X=$55 σ=42%",
    "P3" ~ "Short Call X=$80 σ=30%"
  ))
  

p1 <- greeks_df %>%
  ggplot2::ggplot(aes(x = S, y = delta, color = Position, linetype = Position)) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::geom_vline(xintercept = F0, linetype = "dashed", color = "gold", linewidth = 0.5) +
  ggplot2::annotate("text", x = F0, y = Inf, vjust = 1.5,hjust = 1,label = "Current WTI Price", color = "gold", size = 3)+
  ggplot2::scale_x_continuous(labels = scales::dollar_format()) +
  ggplot2::scale_y_continuous(labels = scales::dollar_format())+
  labs(title = "Delta", x = "Futures Price ($/bbl)", y = "Delta") +
  theme_minimal()

p2 <- greeks_df %>%
  ggplot2::ggplot(aes(x = S, y = gamma, color = Position, linetype = Position)) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::geom_vline(xintercept = F0, linetype = "dashed", color = "gold", linewidth = 0.5) +
  ggplot2::annotate("text", x = F0, y = Inf, vjust = 1.5,hjust = 1,label = "Current WTI Price", color = "gold", size = 3)+
  ggplot2::scale_x_continuous(labels = scales::dollar_format()) +
  ggplot2::scale_y_continuous(labels = scales::dollar_format())+
  labs(title = "Gamma", x = "Futures Price ($/bbl)", y = "Gamma") +
  theme_minimal()

p3 <- greeks_df %>%
  ggplot2::ggplot(aes(x = S, y = vega, color = Position, linetype = Position)) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::geom_vline(xintercept = F0, linetype = "dashed", color = "gold", linewidth = 0.5) +
  ggplot2::annotate("text", x = F0, y = Inf, vjust = 1.5,hjust = 1,label = "Current WTI Price", color = "gold", size = 3)+
  ggplot2::scale_x_continuous(labels = scales::dollar_format()) +
  ggplot2::scale_y_continuous(labels = scales::dollar_format())+
  labs(title = "Vega", x = "Futures Price ($/bbl)", y = "Vega") +
  theme_minimal()

p4 <- greeks_df %>%
  ggplot2::ggplot(aes(x = S, y = theta, color = Position, linetype = Position)) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::geom_vline(xintercept = F0, linetype = "dashed", color = "gold", linewidth = 0.5) +
  ggplot2::annotate("text", x = F0, y = Inf, vjust = 1.5,hjust = 1,label = "Current WTI Price", color = "gold", size = 3)+
  ggplot2::scale_x_continuous(labels = scales::dollar_format()) +
  ggplot2::scale_y_continuous(labels = scales::dollar_format())+
  labs(title = "Theta", x = "Futures Price ($/bbl)", y = "Theta") +
  theme_minimal()

p1 
p2  
p3
p4

```




```{r q1.d, echo=FALSE}

# A more valuable put means price is dropping so a short put will decrease portfolio value and a long put will increase it as prices fall. 
# A more valuable call means price rallies. So as price increases, a short call decreases portfolio value.

# Need to add gamma here and scale everything by # of barrels as well



port <- t1 %>% 
  dplyr::select(Position,Type,sigma,X) %>% 
  tidyr::crossing(S = F.grid) %>% 
  dplyr::rowwise() %>% 
  dplyr::mutate(price = RTL::GBSOption(S = S, X = X, T2M = T2M, r = r, b = 0, sigma = sigma, type = Type)$price,
                gamma = RTL::GBSOption(S = S, X = X, T2M = T2M, r = r, b = 0, sigma = sigma, type = Type)$gamma,
                vega = RTL::GBSOption(S = S, X = X, T2M = T2M, r = r, b = 0, sigma = sigma, type = Type)$vega,
                theta = RTL::GBSOption(S = S, X = X, T2M = T2M, r = r, b = 0, sigma = sigma, type = Type)$theta,
                delta = RTL::GBSOption(S = S, X = X, T2M = T2M, r = r, b = 0, sigma = sigma, type = Type)$delta) %>% 
  dplyr::ungroup()


port_gamma <- port %>% 
  dplyr::select(Position, S, gamma) %>%
  tidyr::pivot_wider(names_from = Position,
                     values_from = gamma,
                     names_prefix = "g") %>% 
  dplyr::mutate(port_gamma = (gP1 * 2000 - gP2 * 2000 - gP3 * 2000)*mult) %>% 
  dplyr::select(S,port_gamma)
  
  
  
  port_summary <- port %>% 
  dplyr::select(Position, S, price) %>%
  tidyr::pivot_wider(names_from = Position,
                     values_from = price,
                     names_prefix = "v") %>% 
  dplyr::mutate(port_value = vP1 * 2000 * mult - vP2 * 2000 * mult - vP3 * 2000 * mult) %>% 
  dplyr::select(S,port_value) %>% 
  dplyr::left_join(port_gamma)
  

(p5 <- port_summary %>% 
    ggplot2::ggplot(aes(x = S))+
    ggplot2::geom_line(aes(y = port_value),color = "blue")+
    geom_vline(xintercept = F0, linetype = "dashed", color = "gold", linewidth = 0.5) +
    ggplot2::annotate("text", x = F0, y = Inf, vjust = 1.5,hjust = 1,label = "Current WTI Price", color = "gold", size = 3)+
    geom_hline(yintercept = 0,linetype = "dashed", color = "darkgrey", linewidth = 0.5)+
    scale_x_continuous(labels = scales::dollar_format()) +
  scale_y_continuous(labels = scales::dollar_format())+
    labs(title = "Portfolio Value as a Function of F", x = "Futures Price ($/bbl)", y = "Portfolio Value",
         subtitle = "Shows that we have virtually unlimited downside risk because of the uncovered short call") +
    theme_minimal())

(p6 <- port_summary %>% 
    ggplot2::ggplot(aes(x = S))+
    ggplot2::geom_line(aes(y = port_gamma),color = "blue")+
    geom_vline(xintercept = 70, linetype = "dashed", color = "gold", linewidth = 0.5) +
    ggplot2::annotate("text", x = 70, y = Inf, vjust = 1.5,hjust = 1,label = "Long Put", color = "gold", size = 3)+
    geom_vline(xintercept = 55, linetype = "dashed", color = "gold", linewidth = 0.5) +
    ggplot2::annotate("text", x = 55, y = Inf, vjust = 1.5,hjust = 1,label = "Short Put", color = "gold", size = 3)+
    geom_vline(xintercept = 80, linetype = "dashed", color = "gold", linewidth = 0.5) +
    ggplot2::annotate("text", x = 80, y = Inf, vjust = 1.5,hjust = 1,label = "Short Call", color = "gold", size = 3)+
    scale_x_continuous(labels = scales::dollar_format()) +
    labs(title = "Portfolio Gamma as a Function of F", x = "Futures Price ($/bbl)", y = "Portfolio GBS Gamma",
         subtitle = "Dashed lines are the strike prices of each position") +
    theme_minimal())





```
 
 + The MTM portfolio value shows us very clearly through its negative slope that we are currently holding a negative delta portfolio. It also shows that the rate of change of delta accelerates as the price moves above $80, this confirms that the short call position is driving most of the risk and negative delta in this portfolio. To further prove that, the short put that has a strike of 55, is OTM for the majority of the chart so it's delta does not contribute heavily to the portfolio even though it has a positive delta. 
 
+ The portfolio gamma is highest near the `$`65 range and lowest at the `$`50 range. The high gamma range corresponds to the long put that we hold, because its strike is at `$`70, as the price moves down to 70, the put gets closer to being ITM and so its delta is nearing -1 and gamma increases accordingly. As for the global minimum at 55, the long put moves deep ITM and so its gamma decays as it moves deeper ITM, because the short put nears the money at 55, its gamma contributes heavily to the portfolio as the call option is deep OTM and the short option is deep ITM, meaning that the short put pulls the entire portfolio gamma down with little resistance from the other positions. 

+ For a delta-hedger, gamma tells them how delta changes with the change in price. So a gamma of -10 means that for a $1 change in the underlying, the delta decreases by 10. Gamma is the basis of re-hedging the portfolio with price moves to remain delta-neutral




# Question 2 (3 points)

Sum Greeks across positions to show **portfolio-level Greeks** at the current market. Present in a `gt::gt()` table. At approximately what futures price does the portfolio delta change sign?

```{r q2, echo=FALSE}

port_delta <- port %>% 
  dplyr::select(Position, S, delta) %>%
  tidyr::pivot_wider(names_from = Position,
                     values_from = delta,
                     names_prefix = "d") %>% 
  dplyr::mutate(port_delta = (dP1 * 2000 - dP2 * 2000 - dP3 * 2000)*mult) %>% 
  dplyr::select(S,port_delta)

port_vega <- port %>% 
  dplyr::select(Position, S, vega) %>%
  tidyr::pivot_wider(names_from = Position,
                     values_from = vega,
                     names_prefix = "v") %>% 
  dplyr::mutate(port_vega = (vP1 * 2000 - vP2 * 2000 - vP3 * 2000)*mult) %>% 
  dplyr::select(S,port_vega)

port_theta <- port %>% 
  dplyr::select(Position, S, theta) %>%
  tidyr::pivot_wider(names_from = Position,
                     values_from = theta,
                     names_prefix = "t") %>% 
  dplyr::mutate(port_theta = (tP1 * 2000 - tP2 * 2000 - tP3 * 2000)*mult) %>% 
  dplyr::select(S,port_theta)

(port_greeks <- port_delta %>% 
  dplyr::left_join(port_gamma) %>% 
  dplyr::left_join(port_vega) %>% 
  dplyr::left_join(port_theta) %>% 
  gt::gt() %>% 
  gt::tab_header(
    title    = "WTI Portfolio Greeks") %>% 
  gt::fmt_number(decimals = 0)) %>% 
  gt::cols_align(align = "center")



```


+ Because the portfolio has three positions of 2000 options each, and because they are on the same underlying, the portfolio will never have a positive delta. The reason is that the short call and long puts both have negative deltas so the short put is the only position with a positive delta and it is not large enough to make the portfolio delta positive. 


# Question 3 (8 points)

Overnight, US-Iran tensions escalate sharply. The US strikes Iranian nuclear facilities; Iran retaliates by mining the Strait of Hormuz. WTI gaps up at the open:

+ WTI futures: F moves from **73.00** to **90.00** (+$17.00/bbl)
+ Implied volatility: each position's sigma spikes by **+0.20** (20 vol points)
+ Time: **1 calendar day** passes

For each position and for the portfolio total, compute:

1. **Actual P&L**: full revaluation at new F, new sigma, new T. P&L = (new price − old price) × n × mult.
2. **Attributed P&L**: ΔV ≈ (Delta × ΔF) + (½ × Gamma × ΔF²) + (Vega × Δσ) + (Vanna × ΔF × Δσ) + (Theta × Δt), scaled to dollars.
3. **Unexplained residual**: Actual − Attributed.
4. **Attribution breakdown**: what share (%) of Attributed P&L comes from each Greek?

**Vanna** (∂²V/∂F∂σ) measures how delta changes as volatility moves. With both a large spot jump and a large vol spike coinciding, Vanna is non-negligible. Use a central-difference cross-partial:

$$\text{Vanna} = \frac{V(F+h, \sigma+k) - V(F+h, \sigma-k) - V(F-h, \sigma+k) + V(F-h, \sigma-k)}{4 \cdot h \cdot k}$$

with h = `step_S` and k = `step_v`. The Vanna P&L contribution is Vanna × ΔF × Δσ.

Present all results in a `gt::gt()` table. Which position dominates the loss? Is the Greek attribution a reliable approximation here, and why or why not?

```{r q3, echo=FALSE}

# Define Variables
f_jump <- 17
vol_jump <- 0.2
t <- 1/365


# Calculate Vanna First




iran_price <- t1 %>%
  dplyr::select(Position, Type, X, sigma, delta, gamma,
                vega, theta, P_current, n_contracts) %>%
  dplyr::mutate(
    P_actual = purrr::pmap_dbl(
      .l = list(S=F0+f_jump, X=X, T2M=T2M-t, r=r, b=0, N=1000,
                sigma=sigma+vol_jump, type=Type, optionStyle="american"),
      .f = calculate_crr),
    vanna_pp = purrr::pmap_dbl(
      .l = list(S=F0+delta_step, X=X, T2M=T2M, r=r, b=0, N=1000,
                sigma=sigma+vega_step, type=Type, optionStyle="american"),
      .f = calculate_crr),
    vanna_pm = purrr::pmap_dbl(
      .l = list(S=F0+delta_step, X=X, T2M=T2M, r=r, b=0, N=1000,
                sigma=sigma-vega_step, type=Type, optionStyle="american"),
      .f = calculate_crr),
    vanna_mp = purrr::pmap_dbl(
      .l = list(S=F0-delta_step, X=X, T2M=T2M, r=r, b=0, N=1000,
                sigma=sigma+vega_step, type=Type, optionStyle="american"),
      .f = calculate_crr),
    vanna_mm = purrr::pmap_dbl(
      .l = list(S=F0-delta_step, X=X, T2M=T2M, r=r, b=0, N=1000,
                sigma=sigma-vega_step, type=Type, optionStyle="american"),
      .f = calculate_crr),
    vanna    = ((vanna_pp - vanna_pm - vanna_mp + vanna_mm) /
               (4 * delta_step * vega_step)) * n_contracts * mult,
    actual_PL = (P_actual - P_current) * n_contracts * mult,
    PL_delta  = delta  * f_jump,
    PL_gamma  = 0.5    * gamma  * f_jump^2,
    PL_vega   = vega   * vol_jump,
    PL_theta  = theta,
    PL_vanna  = vanna  * f_jump * vol_jump,
    attr_PL   = PL_delta + PL_gamma + PL_vega + PL_theta + PL_vanna,
    residual  = actual_PL - attr_PL,
    resid_pct = residual / abs(actual_PL) * 100) %>% 
  dplyr::mutate(
    pct_delta = PL_delta / attr_PL * 100,
    pct_gamma = PL_gamma / attr_PL * 100,
    pct_vega  = PL_vega  / attr_PL * 100,
    pct_theta = PL_theta / attr_PL * 100,
    pct_vanna = PL_vanna / attr_PL * 100
  ) %>%
  dplyr::select(-vanna_pp, -vanna_pm, -vanna_mp, -vanna_mm)


iran_price %>%
  dplyr::select(Position, actual_PL, attr_PL, residual, resid_pct,
                PL_delta, PL_gamma, PL_vega, PL_theta, PL_vanna,
                pct_delta, pct_gamma, pct_vega, pct_theta, pct_vanna) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = gt::md("**P&L Attribution — WTI Options Portfolio**")) %>% 
  gt::tab_spanner(label = "P&L ($)",
                  columns = c(actual_PL, attr_PL, residual, resid_pct)) %>%
  gt::tab_spanner(label = "Attribution by Greek ($)",
                  columns = c(PL_delta, PL_gamma, PL_vega, PL_theta, PL_vanna)) %>%
  gt::tab_spanner(label = "% of Attributed P&L",
                  columns = c(pct_delta, pct_gamma, pct_vega, pct_theta, pct_vanna)) %>% 
  gt::cols_label(
    Position  = "Position",
    actual_PL = "Actual",
    attr_PL   = "Attributed",
    residual  = "Residual ($)",
    resid_pct = "Residual (%)",
    PL_delta  = "Delta",  
    PL_gamma = "Gamma",
    PL_vega   = "Vega",  
    PL_theta = "Theta",  
    PL_vanna = "Vanna",
    pct_delta = "Delta",  
    pct_gamma = "Gamma",
    pct_vega  = "Vega",  
    pct_theta = "Theta", pct_vanna = "Vanna") %>% 
  gt::fmt_currency(
    columns  = c(actual_PL, attr_PL, residual,
                 PL_delta, PL_gamma, PL_vega, PL_theta, PL_vanna),
    decimals = 0
  ) %>%
  gt::fmt_number(
    columns  = c(resid_pct, pct_delta, pct_gamma,
                 pct_vega, pct_theta, pct_vanna),
    decimals = 1,
    suffix   = "%"
  ) %>% 
  gt::cols_align(align = "auto")



```


+ Like we noted earlier, the short call option is the position that carries most of the risk and is the position that is driving most of the losses in this scenario. While the long put position also loses, it is not as nearly close to the amount that the call loses. The short put gains because the option holder will not exercise under this scenario so we can keep the option premium without having to deliver on our obligations. 

+ Using the greeks to approximate the P&L attribution is not very accurate because the numerical greeks were calculated when the futures price was at $73. At that price point, the long put & short call positions were both near the money, meaning that they had higher absolute deltas and vegas which is not the case after the shift in price and volatility. 


# Question 4 (4 points)

The Iran war scenario (ΔF = +$17) is an extreme move. How well does the Greek approximation hold as WTI moves further from the current market?

Holding vol and time constant at their initial values, compute Actual P&L (full revaluation) and Attributed P&L (Delta + Gamma only) for the portfolio across ΔF ∈ [–20, +20] in $1 increments.

**a.** Plot Actual, Attributed, and the Residual on the same chart. Mark the Iran war move (ΔF = +17) with a vertical line.

**b.** At what ΔF does the absolute residual first exceed 5% of the absolute Actual P&L? Report in a `gt::gt()` table. What does this tell you about using Greeks for risk management in extreme scenarios?

```{r q4, echo=FALSE}

price_grid <- seq(from = -20, to = 20, by = 1)

test <- t1 %>% 
  dplyr::select(Position,Type,sigma,X,n_contracts,P_current,delta,gamma) %>% 
  tidyr::crossing(price_grid) %>% 
  dplyr::rowwise() %>% 
  dplyr::mutate(actual_price = pmap_dbl(
    .l = list(S=F0 + price_grid, X=X, T2M=T2M, r=r, b=0, N=1000,
                sigma=sigma, type=Type, optionStyle="american"),
    .f = calculate_crr
  ),
  actual_PL = (actual_price - P_current) * n_contracts * mult,
  attribute_PL = (delta * price_grid) + (0.5 * gamma * price_grid^2),
  residual = actual_PL - attribute_PL) %>% 
  dplyr::ungroup()
  

(p7 <- test %>%
  dplyr::group_by(price_grid) %>%
  dplyr::summarise(
    Actual     = sum(actual_PL),
    Attributed = sum(attribute_PL),
    Residual   = sum(residual),
    .groups    = "drop") %>%
  tidyr::pivot_longer(
    cols      = c(Actual, Attributed, Residual),
    names_to  = "Series",
    values_to = "Value") %>%
  ggplot2::ggplot(aes(x = price_grid, y = Value, color = Series,linetype = Series)) +
  ggplot2::geom_line() +
  ggplot2::geom_vline(xintercept = 17, color = "red",linetype = "dashed",linewidth  = 0.6) +
  ggplot2::annotate("text", x = 17,y = Inf,vjust = 1.5,hjust = 1.2,label = "Iran War Shock (+$17)",color = "red",size    = 3) +
  ggplot2::scale_linetype_manual(
    values = c("Actual" = "solid",
               "Attributed" = "solid",
               "Residual" = "dashed")) +
  ggplot2::scale_x_continuous(labels = scales::dollar_format()) +
  ggplot2::scale_y_continuous(labels = scales::dollar_format()) +
  ggplot2::labs(
    title    = "Actual vs Attributed P&L",
    subtitle = "Greeks approximate actual P&L closely for small moves — divergence grows with shock size",
    x        = "Futures Price Change ($/bbl)",
    y        = "Portfolio P&L ($)") +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom"))



test %>%
  dplyr::group_by(price_grid) %>%
  dplyr::summarise(
    Actual     = sum(actual_PL),
    Residual   = sum(residual),
    .groups    = "drop") %>%
  dplyr::mutate(Residual_pct = abs(Residual) / abs(Actual)) %>% 
  replace_na(list(Residual_pct = 0)) %>% 
  gt::gt() %>% 
  gt::cols_label(Residual_pct = "Residual % of Actual") %>% 
  gt::fmt_currency(columns = c("Actual","Residual"), decimals = 0) %>% 
  gt::fmt_percent(columns = Residual_pct,decimals = 2) %>% 
  gt::cols_align(align = "center",
                 columns = everything()) %>% 
  gt::tab_style(
    style = gt::cell_fill(color = "gold",alpha = 0.5),
    locations = gt::cells_body(
    rows = price_grid %in% c(-2, 2)
    )
  )
  

  


```



+ It is obvious that using static greeks for risk hedging is unfeasible in extreme scenarios, as shown in the table above, the residual value exceeds 5% of the actual P&L when the futures price moves by only $2 in either direction. When it moves by 17, like what happened recently, the greeks mis-attributed the P&L by about 50%.

# Question 5 (6 points)

Discuss the key learnings and insights from a Greek risk management perspective. Your explanation must:

+ Interpret the **full revaluation MTM profile** from Q1(d): describe the portfolio's curvature structure across the $50–$100 range. Identify the regions of maximum and minimum Gamma. Explain what the changing Gamma means for a desk running this portfolio — specifically, how frequently delta rebalancing is needed as WTI moves through the strikes.
+ Interpret the **full revaluation P&L and Gamma profiles** from Q1(d): describe the portfolio's curvature structure and identify which positions drive positive vs negative Gamma regions.
+ Identify which position and which Greek drove the largest loss in the Iran war scenario, and explain why that was predictable from the portfolio structure.
+ Explain the role of **Vanna**: why is it non-trivial when both a large spot jump and a large vol spike occur simultaneously?
+ Explain why the Greek approximation (Q4) breaks down at extreme moves, and what this means for a real desk managing this book.

Pay particular attention to why textbook delta-hedging assumptions break down in geopolitical gap scenarios, and what risk tools a real energy desk would use instead.

*Your answer here.*
